Learning Libraries of Subroutines for Neurally--Guided Bayesian Program Learning

### 2018-05-18 nips_2018.tex
Fiddling and manually merging

%% \begin{table}[t]
%%   \centering
%%     \begin{tabular}{l}
%%       \toprule
%%       $f_0($\code{s}$,$\code{a}$,$\code{b}$) \,=\, $\code{(map ($\lambda$ (x) (if (= x a) b x)) s)}\\
%%       \hspace{0.5cm}($f_0$: \emph{Performs character substitution)}\\
%%       $f_1($\code{s}$,$\code{c}$) \,=\, $\code{(foldr s s ($\lambda$ (x a) (cdr (if (= c x) s a))))}\\
%%         \hspace{0.5cm}($f_1$: \emph{Drop first characters from }\code{s}\emph{ until  }\code{c}\emph{ reached})\\
%%       $f_2($\code{s}$) \,=\, $\code{(unfold s is-nil car ($\lambda$ (z) (}$f_1$\code{ z SPACE)))}\\
%%         \hspace{0.5cm}($f_2$: \emph{Abbreviates a sequence of words. See Fig.~\ref{initialExampleDSL}})\\
%%         $f_3($\code{s}$,$\code{c}$) \,=\, $\code{(foldr s s ($\lambda$ (x a) (if (= c x) nil (cons x a))))}\\
%%         \hspace{0.5cm}($f_3$: \emph{Take characters from }\code{s}\emph{ until }\code{c}\emph{ reached)}
%%         %($\lambda$ ($\lambda$ (foldr $0 $0 ($\lambda$ ($\lambda$ (if (char-eq? $1 $3) nil (cons $1 $0)))))))
%%   \\\bottomrule
%%     \end{tabular}
%%   \caption{Some text editing learned subroutines }\label{textPrimitives}
%%   \end{table}

%% \begin{wrapfigure}{r}{0.6\textwidth}
%%   \centering\begin{tabular}{c}\toprule
%% +106 769-858-438 $\longrightarrow$106.769.858.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
%% +83 973-757-831 $\longrightarrow$83.973.757.831%& Andrew Cencici $\longrightarrow$ Dr. Andrew
%% %DkI$>$DwKI$<$xaPoc& 	DWKI\\
%% %KfvTh$>$Dbl$<$OIJso 	&DBL
%% \\\midrule
%% %(lambda (#(lambda (lambda (map (lambda (if (char-eq? $0 $1) $2 $0))))) '.' SPACE (#(lambda (lambda (map (lambda (if (char-eq? $0 $1) $2 $0))))) '.' '-' (cdr $0))))
%% \multicolumn{1}{c}{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  '.' ' ' ($f_0$ '.' '-' (cdr s)))}}
%% %\multicolumn{1}{l}{$f(\text{\code{s}}) = $\code{(}$f_0$\code{ "Dr. " (}$f_3$\code{ s " "))}}
%% \\
%% \bottomrule
%%   \end{tabular}

%%   %% \vspace{0.25cm}

%%   %% \begin{minipage}{\linewidth}
%%   %%   \centering
%%   %%   $f($\code{s}$) \,=\, $
%%   %% \end{minipage}

%%   \caption{String edit task (top) and the program \system writes for it (bottom). $f_0$ is a subroutine written by \systemEnding, defined in Tbl.~\ref{textPrimitives}.}\label{exampleTextProblem}
%% \end{wrapfigure}

@@

%% \begin{align*}
%%   \log \probability[x|p] \triangleq& \sum_{n\leq N}\log
%%   \mathcal{N}(p(i_n,\vec{\mathcal{R}}^*)|o_n) - \frac{D\log N}{2}
%% \end{align*}

%% %% and integrate over the real-valued parameters, which we collectively write as $\vec{\mathcal{R}}$:
%% %% \begin{align*}
%% %% \log   \probability\left[\{(i_n,o_n)\}|p\right] = \log \int \mathrm{d}\vec{\mathcal{R}}\; P(\vec{\mathcal{R}})\prod_{n\leq N}\mathcal{N}(p(i_n,\vec{\mathcal{R}})|o_n)
%% %% %\text{\emph{BIC:}} &\approx 
%% %% \end{align*}
%% where $\mathcal{N}(\cdot |\cdot )$ is the normal density and
%% $P_{\vec{\mathcal{R}}}(\cdot )$ is a prior over
%% $\vec{\mathcal{R}}$. We approximate this marginal using the BIC~:

%% where $\vec{\mathcal{R}}^*$ is an assignment to $\vec{\mathcal{R}}$
%% found by performing gradient ascent on the likelihood of the
%% observations w.r.t. $\vec{\mathcal{R}}$.

\begin{wrapfigure}{r}{0.45\textwidth}\vspace{-0.5cm} \newcommand{\functionSize}{1.5cm}\centering
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \includegraphics[width = \functionSize]{figures/functions/181.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/149.png}
    \includegraphics[width = \functionSize]{figures/functions/187.png}
  \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top row) and polynomials (bottom row) while the recognition  model jointly learns to look at a graph of the function (above) and predict which of those subroutines best explains the observation.}\label{functions}\vspace{-2cm}
    \end{wrapfigure}\system learns a DSL containing 13 new functions,
most of which are templates for polynomials of different orders or ratios of polynomials.
It also learns to find programs that minimize the number of continuous degrees of freedom.
For example, it learns to represent linear functions with the program
\code{(* real (+ x real))}}, which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing -- both the implicit bias towards shorter programs and the likelihood model's BIC penalty.

@@

We compare with ablations of our model on held out tasks.
The purpose of this ablation study is 
both to examine the role of each component of \systemEnding,
as well as to compare with
prior approaches in the literature:
A head-to-head
comparison of program synthesizers is complicated by the fact that
each system, including ours, makes idiosyncratic 
assumptions about the space of programs and the statement of tasks.
Nevertheless, much prior work can be modeled as
special cases of our setup. We compare with the following baselines:
\\\noindent \textbf{No NN}, which lesions the recognition model.
\\\noindent \textbf{NPS}, which does not learn the DSL,
instead learning the recognition model
from samples drawn from the fixed DSL.
We call this NPS (Neural Program Synthesis)
because this setup is closest to how
RobustFill~\cite{devlin2017robustfill} and DeepCoder~\cite{balog2016deepcoder} are trained.
\\\noindent \textbf{SE}, which lesions the recognition model and restricts the DSL  learning algorithm to
only add subexpressions of programs in the frontiers to the DSL. This is how prior approaches have learned libraries of functions~\cite{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,DBLP:conf/ecai/LinDETM14}.
\\\noindent \textbf{Enum}, which enumerates a frontier without any learning -- equivalently, our first search step.

@@

 \section{Contribution and Outlook}

\parbox{0.58\textwidth}{
We contribute an algorithm, \systemEnding, that learns to program by bootstrapping a DSL with new
domain-specific primitives that the algorithm itself discovers,
together with a neural recognition model that learns how to
efficiently deploy the DSL on new tasks. We believe this integration
of top-down symbolic representations and bottom-up neural networks --
both of them learned -- could help make program induction systems more
generally useful for AI. Many directions remain open. Two immediate
goals 
are to integrate more sophisticated neural recognition
models~\cite{devlin2017robustfill} and program
synthesizers~\cite{solar2008program}, which may improve performance in
some domains over the generic methods used here.
Another direction is to
explore DSL meta-learning: Can we find a \emph{single} universal
primitive set that could effectively bootstrap DSLs for new domains,
including the three domains considered,  but also many others?
}\hfill
\begin{minipage}{0.4\textwidth}
  \centering
  \includegraphics[width=\textwidth]{figures/geomCompiled.eps} 
  \captionof{figure}{
  Top left: a few example of the shapes used as targets. Top right: a few
  examples of discovered new shapes. Bottom: learning curve across iterations in
  percent in the middle, and on both side average of the compiled new shapes
  before any traning and at the end of the last iteration. See supplementary
  material for more details.}
\label{geomCompiled}
\end{minipage}

>>>>

@@

\begin{wrapfigure}{r}{0.45\textwidth}\vspace{-0.5cm} \newcommand{\functionSize}{1.5cm}\centering
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}\includegraphics[width = \functionSize]{figures/functions/181.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/149.png}
    \includegraphics[width = \functionSize]{figures/functions/187.png}
  \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top row) and polynomials (bottom row) while the recognition  model jointly learns to look at a graph of the function (above) and predict which of those subroutines best explains the observation.}\label{functions}\vspace{-2cm}
\end{wrapfigure}
\system learns a DSL containing 13 new functions,
most of which are templates for polynomials of different orders or ratios of polynomials.
It also learns to find programs that minimize the number of continuous degrees of freedom.
For example, it learns to represent linear functions with the program
\code{(* real (+ x real))}}, which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing -- both the implicit bias towards shorter programs and the likelihood model's BIC penalty.

@@

We compare with ablations of our model on held out tasks.
The purpose of this ablation study is 
both to examine the role of each component of \systemEnding,
as well as to compare with
prior approaches in the literature:
a head-to-head
comparison of program synthesizers is complicated by the fact that
each system, including ours, makes idiosyncratic 
assumptions about the space of programs and the statement of tasks.
\begin{wraptable}{r}{0.55\textwidth}
\vspace{-5pt}
%% This radically shrinks the table
\tabcolsep=2.5pt
\renewcommand{\arraystretch}{0.5}
\begin{tabular}{lcccccc}
  \toprule& Ours& 
    No NN & SE&NPS & PCFG & Enum \\
  \midrule
  \multicolumn{7}{c}{\emph{List Processing}}\\
  \midrule
  \% solved&\textbf{79\%} &76\% &71\%&35\%&62\%&37\%\\
  Solve time&  4.1s&5.8s&10.6s&34.7s&43.4s&20.2s\\
  \midrule
  \multicolumn{7}{c}{\emph{Text Editing}}\\
  \midrule
  \% solved&\textbf{74\%} &43\% &30\%&33\%&0\%&4\%\\
  Solve time& 38s&65s&38s&80s&--&235s\\
  \midrule
  \multicolumn{7}{c}{\emph{Symbolic Regression}}\\
  \midrule
  \% solved&   \textbf{84\% }&75\%&62\%&38\%&38\%&37\% \\
  Solve time&  \textbf{24}s& 40s  &28s&31s&55s&29s\\
  \bottomrule
  \end{tabular}
\caption{\% solved before timeout. Solve time: averaged over solved
  tasks.  SE: NPS: trained like RobustFill/DeepCoder.  PCFG: model w/o
  structure learning.  Enum: model w/o any learning.  %% MDL:
  %% $-\expect\left[\probability[x|p] \right]$. For domains other than
  %% symbolic regression MDL is 0 nats.
  %% For symbolic regression MDL is a
  %% proxy for \# of continuous parameters.
}\vspace{-0.5cm}\label{baselineComparisons} \end{wraptable}
Nevertheless, much prior work can be modeled as
special cases of our setup. 
We compare with the following baselines:
\\\noindent \textbf{No NN}, which lesions the recognition model.
\\\noindent \textbf{NPS}, which does not learn the DSL,
instead learning the recognition model
from samples drawn from the fixed DSL.
We call this NPS (Neural Program Synthesis)
because this setup is closest to how
RobustFill~\cite{devlin2017robustfill} and DeepCoder~\cite{balog2016deepcoder} are trained.
\\\noindent \textbf{SE}, which lesions the recognition model and restricts the DSL  learning algorithm to
only add subexpressions of programs in the frontiers to the DSL. This is how prior approaches have learned libraries of functions~\cite{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,DBLP:conf/ecai/LinDETM14}.
\\\noindent \textbf{Enum}, which enumerates a frontier without any learning -- equivalently, our first search step.

@@

\section{Discussion}

\parbox{0.58\textwidth}{

\paragraph{Outlook.}
We contribute an algorithm, \systemEnding, that learns to program by
bootstrapping a DSL with new domain-specific primitives that the algorithm
itself discovers, together with a neural recognition model that learns how to
efficiently deploy the DSL on new tasks. We believe this integration of top-down
symbolic representations and bottom-up neural networks --- both of them learned
--- helps make program induction systems more generally useful for AI. Many
directions remain open.

\paragraph{Future.}
Two immediate goals are to integrate more sophisticated neural recognition
models~\cite{devlin2017robustfill} and program
synthesizers~\cite{solar2008program}, which may improve performance in some
domains over the generic methods used here. Figure \ref{geomCompiled} gives an
example of a promising domain for this, abstract geometry, see supplementary material for details.  Another direction is to explore DSL
meta-learning: can we find a \emph{single} universal primitive set that could
effectively bootstrap DSLs for new domains, including the three domains
considered,  but also many others?
}\hfill
\begin{minipage}{0.4\textwidth}
  \centering
  \includegraphics[width=\textwidth]{figures/geomCompiled.eps} 
  \captionof{figure}{
  Top left: a few example of the shapes used as targets. Top right: a few
  examples of discovered new shapes. Bottom: learning curve across iterations in
  percent in the middle, and on both side average of the compiled new shapes
  before any traning and at the end of the last iteration.}
\label{geomCompiled}
\end{minipage}
